\MOC Server Pages:\\Web Publishing as Staged Computation

### 2007-04-16 paper/paper.tex
svn-r28: 1st journal revision
darcs-hash:20070416223336-e5a07-01112a05bd1eb888105e8ace6bcd3740ef742de7.gz

\documentclass[preprint]{acm_proc_article-sp}
\usepackage{ucs,textcomp,url,xspace,graphicx,color}
\usepackage[nolineno]{lgrind}
\usepackage[bf]{caption2}
\usepackage[numbers]{natbib}
\usepackage[utf8]{inputenc}
\ifx\pdfoutput\undefined\else
  \usepackage[
    pdfpagemode=None,
    pdftitle={MetaOCaml Server Pages},
    pdfsubject={Web Publishing as Staged Computation},
    pdfauthor={Christopher League}
  ]{hyperref}
\fi
\input{fonts}
\def\MOC{MetaOCaml\xspace}
\hyphenation{Meta-OCaml data-base Slash-dot}
\def\quote{\list{}{\leftmargin1.4em\rightmargin1em}\item[]}
\newcommand{\myfig}[3]{%
  \begin{figure}[tbp]%
    #3%
    \caption{#2}%
    \label{fig:#1}%
  \end{figure}}
\newcommand{\screenshot}[2]{%
  \myfig{#1}{#2}
  {\centering%
    \includegraphics[width=20pc,keepaspectratio]{#1}}}
\newcommand{\gnuplot}[2]{%
  \myfig{#1}{#2}{\input{#1}}}
\newcommand{\listing}[2]{\lagrind[tbp]{#1}{#2}{lst:#1}}

\newcounter{listing}
\setcounter{listing}{0}
\makeatletter
\def\ext@listing{lol}
\def\fnum@listing{Listing\nobreakspace\thelisting}
\makeatother
\newcommand{\mylst}[2]{%
  \captionof{listing}{\textbf{#1.} #2}%
  \label{lst:#1}%
  \lgrindfile{#1}}
\newcommand{\mylstf}[2]{%
  \begin{figure}[tbp!]%
    \lgrindfile{#1}%
    \captionof{listing}{\textbf{#1.} #2}%
    \label{lst:#1}%
  \end{figure}}

\title{\MOC Server Pages:}
\subtitle{Web Publishing as Staged Computation}
\numberofauthors{1}
\author{
  \alignauthor Christopher League\\
  \affaddr{Long Island University · Computer Science}\\
  \affaddr{1 University Plaza · Brooklyn, NY 11201}\\
  \email{christopher.league@liu.edu}}
%\conferenceinfo{\MOC Workshop}{October 2004, Vancouver}
%\CopyrightYear{2004}
%\toappear{Submitted to the \MOC Workshop (Vancouver,
%  2004).}

\begin{document}
\maketitle
\begin{abstract}
  Modern dynamic web services are really computer programs.
  Some parts of these programs run off-line, others run
  server-side on each request, and still others run within
  the browser.  In other words, web publishing is
  \emph{staged computation,} either for better performance,
  or because certain resources are available in one stage
  but not another.  Unfortunately, the various web
  programming languages target one stage only.  This is a
  tremendous opportunity for multi-stage languages in
  general, and for \MOC in particular.

  We present the design of \MOC Server Pages.  Unlike
  other languages in its genre, the embedded \MOC code
  blocks may be annotated with staging information, so that
  the programmer may safely and precisely control which
  computation occurs in which stage.  A prototype web
  server, written in OCaml, supports web sites with both
  static and dynamic content.  We provide several sample
  programs and demonstrate the performance gains won using
  multi-stage programming.
\end{abstract}
\category{D.3.2}{Programming Languages}{Language
  Classifications}[Specialized application languages]
\category{H.3.5}{Information Storage and Retrieval}{Online Information
  Services}[Web-based services]

@@

There is at least one dictum of program design that we cannot escape
on the web: \emph{performance matters.}  As a web publisher, visitors
are your bread and butter.  But will your servers and scripts be ready
for the day that your site is featured on prime time television, or on
\texttt{slashdot.org}?  If tens of thousands of potential users drop
by to find a sluggish (or dead) server, most of them will never
return~\cite{greenspun99panda}.

This happens so often to sites featured on Slashdot---a
``news for nerds'' discussion site---that it has come to be
known as the \emph{Slashdot effect:} ``a site that might be
designed to handle a few hundred hits per day can suddenly
find itself handling that many a
second.''~\cite{turner03slashdot}~ Although some ISPs have
bandwidth limitations, Slashdot creator Rob Malda says most
sites that fall suffer from poor planning and architecture:
\begin{quote}
  The problem is, if you have any sort of complicated code
  on your page, that’s what kills people. I generally think
  that sites are not dying because they fill their pipe, but
  because they have poorly written code.

  As a general rule, anybody who has a pretty good
  understanding of web design, they’ve done a good job of
  learning what information to cache. What information needs
  to be pregenerated. So when you’re actually loading a
  page, even if it’s a complicated page that looks dynamic
  and custom, on the back end of that, what they’re really
  doing is putting together a bunch of puzzle pieces that
  have been pregenerated, and making the simplest, quickest
  decisions they possibly can.~\cite{turner03slashdot}
\end{quote}

This quote points (informally) to the idea of \emph{staging} the
computation performed by a web service.  Indeed, web publishing is an
application area that is naturally staged:\footnote{Nevertheless, it
  seems the literature on meta-programming has not yet acknowledged
  web services as a potential application area, although
  \citet{sheard01accomp} does
  mention \emph{mobile code.}} %
\begin{enumerate}
\item Content (text, images, programs, etc.) created
  off-line is uploaded to the server---the \emph{publish}
  stage;

@@

\screenshot{confcal}{A simple conference calendar, implemented as a
  PHP script that connects to a PostgreSQL database.  Not as
  sophisticated as the service described in the text.}

Now, how might this conference service be staged?  Can
anything be computed off-line (at the \emph{publish} stage)?
Yes!  Since this page is probably part of a much larger site
(whose structure does not change every day), the menus and
other navigation aids can be laid out in advance.  We will
not know which conferences to display until the user
presents some identification (in the form of a `cookie'),
but since the conference data changes infrequently, it may
help to prepare the text of each row in advance.

@@

At this point, we should emphasize the importance of
\emph{profiling} in developing scalable web services.  This
particular design for the conference calendar may not be
optimal, depending on the relative speeds of the CPU,
memory, database, network, and disk.  Rather, our aim is to
provide a single language in which the various staging
possibilities can be expressed naturally.

This approach is in stark contrast to the status quo, where each
system targets one stage only.  One such system, the Website Meta
Language~\cite{engelschall-wml}, is an ``off-line HTML generation
toolkit'' designed for the \emph{publish} stage.  But many other
programs (and countless ad-hoc scripts) spit out HTML pages:
\LaTeX2HTML, for example.  Google reports surprisingly many
programs\footnote{\url{http://google.com/search?q=gedcom+generate+html}}
for creating family tree web sites from genealogy database files;
these also count as publish-stage tools.

The \emph{serve} stage is well-served by the ``server page''
languages, including JSP~\cite{sun-jsp},
ASP~\cite{microsoft-asp},
MSP~\cite{elsman03web}, and
PHP~\cite{bakken04php}.  The CGI protocol~\cite{w3c-cgi}
addresses the \emph{serve} stage, as do the embedded
interpreters (such as \texttt{mod\_perl} and
\texttt{mod\_python} for Apache) that exist to ameliorate some
of the overhead of CGI.

There is, relatively, a paucity of languages that operate
client-side (\emph{display} stage), probably due to the
difficulty of securing an installed base of interpreters.
JavaScript~\cite{netscape-javascript} and Java applets are
notable exceptions.

Imagine implementing the conference calendar, as conceived
above, using currently deployed technology: a Perl script
outputs a PHP page which contains JavaScript!  Values are
passed from one stage to the next as strings, and the
programmer must manage all the quoting and persistence
issues by hand.

We present ``\MOC server pages,'' a new domain-specific language for
web applications programming.  It leverages the staging annotations of
\MOC~\cite{calcagno03meta,taha00metaml} to provide safe and precise
control over the first two stages.  (We leave further consideration of
the display stage as future work.)  The system is implemented as two
components: a translator transforms the server page language into a
\MOC module, which then can be incorporated into our multi-threaded
HTTP/1.1 server (also written in \MOC).  The scalability gained by
staging certain applications is stunning.

The next section sketches the design and translation of \MOC
server pages, and section~\ref{sec:eg} includes some non-trivial
examples.  The server implementation is decribed in
section~\ref{sec:impl}.  Performance and scalability are
discussed in section~\ref{sec:perf}

@@

The general idea of a \emph{server page} language is that we
  write HTML by default, and embed code %
\cd{\Mquote{}like this\Munquote}.
PHP programmers are familiar with this syntax for embedding code, but
in our case, the code itself is written in \MOC.  Here is your first
\MOC server page:
\lgrindfile{first.meta}
and its output:
\begin{quote}
  \textit{Hello,} \textbf{Unhandled exception: Failure("Nice try!")}
\end{quote}
If a code block raises an exception, the message is sent to the
client's browser in bold-face, and the rest of the page is
aborted.

A very common use of code blocks is to print out (i.e., send to
the browser) the result of evaluating some expression.  The
syntax \cd{\Mquote= e\Munquote} is designated for this task; \cd{e} must
have type \cd{string}.  Alternatively, messages may be formatted
with \cd{printf} by placing the format string immediately after
the code delimiter: \lgrindfile{pi.meta}
\begin{quote}
  ........ $\pi$ ÷ 008 is 0.3927
\end{quote}
In a departure from the standard OCaml syntax for \cd{printf}, the
arguments are comma-delimited.  This way, fewer parentheses are
required when using the escape or lift operators on the arguments.

One more kind of code delimiter is used for \emph{declarations;}
these are lifted to the top of your program, and evaluated during
the publishing stage:
%%
\lgrindfile{perm.meta}
\begin{quote}
Permissions on current directory are 0755
\end{quote}

Once published, this output will never change!  The \cd{stat}
call is executed only once, not on each request.

@@

To compute and splice in a regular (non-code) value, we define a
function \cd{{\KWfont let} lift x = \Mbegincode x\Mendcode}; this
takes any value and turns it into code.  We use it like this:
\cd{\Mbegincode 2 * \Msplice(lift(3+4))\Mendcode} The addition is
performed immediately (because it is escaped), and the result is
spliced into the code, producing \cd{\Mbegincode 2 * 7\Mendcode}.

@@

\subsection{Translation to \MOC}
To see how all this works, let's consider how a \MOC server
page is translated into a proper \MOC program, to be executed
at publish time.  Since the program is executed before any
browser has requested the page, it cannot directly return or
output HTML.  Instead, it will construct and return a \cd{code}
object which is subsequently run on each request (serve stage).
The third (display) stage proposed in section~\ref{sec:intro} is not
supported by this design.
Listing~\ref{lst:trans.meta} shows a sample \MOC server page,
and listing~\ref{lst:trans.ml} contains its translation.

\mylst{trans.meta}{This file demonstrates the various
  kinds of code blocks.}

\mylst{trans.ml}{An automatic translation of the page in
  listing~\ref{lst:trans.meta}.  Declarations have been lifted, and
  everything else appears within \cd{\Mbegincode brackets\Mendcode}}

Declaration blocks have been lifted to the top; any side effects
contained there are executed when the page is published.  The
constructed code follows the line that begins with \cd{{\KWfont let} page}.

The \cd{a b c} represent publish-stage arguments (the names are
specified in the \cd{pragma}), whereas \cd{req} and \cd{puts} are
(fixed) serve-stage arguments.  \cd{req} encapsulates the HTTP request
details, including the headers and query arguments.  \cd{puts} is a
function, provided by the server, to transmit text across the network
to the user's browser.  \cd{arg} is then defined to look up an
argument in the query string of the request; its type is \cd{string\to
  string option}.
Since \cd{req} and \cd{puts} are serve-stage arguments, it is
incorrect to invoke them in the \emph{publish} (first) stage, and indeed
the \MOC type system prevents this.

A few finer points to note: the translator discards newlines that
immediately follow code blocks (otherwise,
listing~\ref{lst:trans.ml} would be dotted with %
\cd{puts}~\verb+"\n"+ statements).  It automatically appends a
semi-colon or the {\KWfont in} keyword to code blocks, as required (see,
for example, the lines containing \cd{statements},
\cd{more\_statements}, and \cd{{\KWfont let} x} in
listing~\ref{lst:trans.ml}).

\subsection{Staged code blocks}
Now that we understand how the server page is assembled into a staged
program, the effects of adding \MOC staging operators to our pages
should be predictable.  Below is a reformulation of the permissions
example, using escape \Msplice{} and \cd{lift} to ensure that
\cd{stat} is called only once.%
\lgrindfile{perm2.meta}%
Still, the \cd{printf} conversion is performed at serve time, so maybe
it is best to pre-generate that as well:%
\lgrindfile{perm3.meta}%
Now, all that remains to execute at serve time is %
\cd{puts "0755"}.  It is also possible to splice in serve-stage
parameters at publish time, such as in this count-down example:%
\lgrindfile{count.meta}%
It removes the loop overhead by expanding to a sequence of 99
\cd{puts} statements.

These tiny examples illustrate some of the common uses of the
escape operator, for which we developed syntactic sugar.  The first
set of code blocks we define use the tilde character to indicate that
some part of the enclosed quote is escaped:

\begin{tabular}{l@{\quad$\leadsto$\quad}l}
\cd{\Mquote{\Mtilde} a \Munquote} &
\cd{\Mquote{} \Msplice( a ) \Munquote}
\\
\cd{\Mquote{\Mtilde=} b \Munquote} &
\cd{\Mquote{=} \Msplice( b ) \Munquote}
\\
\cd{\Mquote{\Mtilde}let x = c \Munquote} &
\cd{\Mquote{} let x = \Msplice( c ) \Munquote}
\\
\cd{\Mquote{\Mtilde}\id{"fmt"} d, e, f\Munquote} &
\cd{\Mquote{}\id{"fmt"} \Msplice(d), \Msplice(e), \Msplice(f)\Munquote}
\\
\end{tabular}

@@

\begin{tabular}{{l@{\quad$\leadsto$\quad}l}}
\cd{\Mquote{\Mbang} a \Munquote} &
\cd{\Mquote{} \Msplice(lift( a )) \Munquote}
\\
\cd{\Mquote{\Mbang=} b \Munquote} &
\cd{\Mquote{=} \Msplice(lift( b )) \Munquote}
\\
\cd{\Mquote{\Mbang}let x = c \Munquote} &
\cd{\Mquote{} let x = \Msplice(lift( c )) \Munquote}
\\
\cd{\Mquote{\Mbang}\id{"fmt"} d, e\Munquote} &
\cd{\Mquote{}\id{"fmt"} \Msplice(lift(d)), \Msplice(lift(e))\Munquote}
\\
\end{tabular}

We will see examples of most of these blocks in section~\ref{sec:eg}.

@@

\MOC server page support publish-stage arguments, as demonstrated by
the identifiers \cd{a b c} in listings~\ref{lst:trans.meta}
and~\ref{lst:trans.ml}.  The programmer specifies the pattern to be
used with a \cd{\Mquote{}pragma args ...\Munquote} declaration, which
may appear anywhere in the code.

@@

To explore the expressiveness of this design, we now look at a
series of web services, organized to comprise a small web
site.\footnote{A demo is available at
  \url{http://comsci.liu.edu/metaserv/}}
To give the services a similar look, we developed a site-wide style
sheet, and an OCaml function to generate a navigation bar from a
hierarchical list of page titles and links.  The following bindings are
available site-wide:

\lgrindfile{navspec-sig.ml}
%%
The functions above may be invoked at any stage.  Naturally, if the
page title includes a dynamic argument, then \cd{preamble} will have
to be delayed until the serve stage.

@@

\screenshot{power127}{Web browser displaying the result of
  7\textsuperscript{127}.  The URI \texttt{/power127} maps to code
  where the early argument (the exponent) has been set to 127, and the
  late argument (\texttt{x}, the base) is taken from the query
  string.}

Judging from its prevalence in the literature, we are certain that
millions of grateful users would subscribe to an online service
capable of computing the \emph{power} function!  The screen shot in
figure~\ref{fig:power127} illustrates how it works.  The navigation
bar shows the exponents for which code has been
pre-generated.\footnote{In the current implementation, it is not
  possible for the user to request other exponents once the server is
  running.  See section~\ref{sec:impl} for an explanation.}  After
selecting the exponent, the user types the base into the form and
presses return.  The result is computed using the arbitrary-precision
\cd{Num} module of the OCaml library.  The complete script appears in
listing~\ref{lst:power.meta}.

\mylst{power.meta}{This implements the staged power function as a
  web service.  The exponent \cd{y} is an early (publish stage)
  argument; but the base \cd{x} is late (serve stage).}

The user's input is converted to a number relatively late in the
script; see the line marked with {(****)}.  If \cd{num\_of\_string}
generates an exception (perhaps because the user typed non-numeric
text into the box), the navigation bar and form will have already been
output before the error message appears.

>>>>

\documentclass{elsart}
\usepackage{textcomp,url,xspace,graphicx,color}
\def\UrlFont{\tt\small}
\usepackage[utf8]{inputenc}
\usepackage[leftno,norules,nolineno]{lgrind}
%\usepackage[bf]{caption2}
\usepackage[numbers]{natbib}
\ifx\pdfoutput\undefined\else
  \usepackage[
    pdfpagemode=None,
    pdftitle={MetaOCaml Server Pages},
    pdfsubject={Web Publishing as Staged Computation},
    pdfauthor={Christopher League}
  ]{hyperref}
\fi
\input{fonts}
\def\MOC{MetaOCaml\xspace}
\hyphenation{Meta-OCaml data-base Slash-dot}
\def\quote{\list{}{\leftmargin1.4em\rightmargin1em}\item[]}
\newcommand{\myfig}[3]{%
  \begin{figure}[tbp]%
    #3%
    \caption{#2}%
    \label{fig:#1}%
  \end{figure}}
\newcommand{\screenshot}[2]{%
  \myfig{#1}{#2}
  {\centering%
    \includegraphics[width=25pc,keepaspectratio]{#1}}}
\newcommand{\gnuplot}[2]{%
  \myfig{#1}{#2}{\centering\input{#1}}}

\LGindent=16pt
\def\lgrindfile#1{
  \par\vskip.5\baselineskip\begingroup
  \begin{lgrind}\input #1\relax
  \end{lgrind}\endgroup}

\newcommand{\mylstf}[2]{%
  \myfig{#1}{\texttt{#1}.  #2}{\LGnuminterval=3\lgrindfile{#1}}}
\newcommand{\mylst}[2]{%
  \myfig{#1}{#2}{\LGnuminterval=3\lgrindfile{#1}}}

\begin{document}
\begin{frontmatter}
\title{\MOC Server Pages:\\Web Publishing as Staged Computation}
\author{Christopher League}
\address{Long Island University \textperiodcentered{} Computer Science\\
  1 University Plaza \textperiodcentered{} Brooklyn, NY 11201\\
  \textup{\texttt{christopher.league@liu.edu}}}

\begin{abstract}
  Modern dynamic web services are really computer programs.  Some
  parts of these programs run off-line, others run server-side on each
  request, and still others run within the browser.  In other words,
  web publishing is \emph{staged computation,} either for better
  performance, or because certain resources are available in one stage
  but not another.  Unfortunately, the various web programming
  languages make it difficult to spread computation over more than one
  stage.  This is a tremendous opportunity for multi-stage languages
  in general, and for \MOC in particular.

  We present the design of \MOC Server Pages.  Unlike other languages
  in its genre, the embedded \MOC code blocks may be annotated with
  staging information, so that the programmer may safely and precisely
  control which computation occurs in which stage.  A prototype web
  server, written in OCaml, supports web sites with both static and
  dynamic content.  We provide several sample programs and demonstrate
  the performance gains won using multi-stage programming.
\end{abstract}
\end{frontmatter}

%\category{D.3.2}{Programming Languages}{Language
%  Classifications}[Specialized application languages]
%\category{H.3.5}{Information Storage and Retrieval}{Online Information
%  Services}[Web-based services]

@@

There is at least one dictum of program design that we cannot escape
on the web: \emph{performance matters.}  As a web publisher, visitors
are your livelihood.  But will your servers and scripts be ready for
the day that your site is featured on prime time television, or on
\texttt{slashdot.org}?  If tens of thousands of potential users drop
by to find a sluggish (or dead) server, most of them will never
return~\cite{greenspun99panda}.

This happens so often to sites featured on Slashdot---a
``news for nerds'' discussion site---that it has come to be
known as the \emph{Slashdot effect:} ``a site that might be
designed to handle a few hundred hits per day can suddenly
find itself handling that many a
second.''~\cite{turner03slashdot}~ Although some ISPs have
bandwidth limitations, Slashdot creator Rob Malda says most
sites that fail suffer from poor planning and architecture:
\begin{quote}
  The problem is, if you have any sort of complicated code
  on your page, that's what kills people. I generally think
  that sites are not dying because they fill their pipe, but
  because they have poorly written code.

  As a general rule, anybody who has a pretty good
  understanding of web design, they've done a good job of
  learning what information to cache. What information needs
  to be pregenerated. So when you're actually loading a
  page, even if it's a complicated page that looks dynamic
  and custom, on the back end of that, what they're really
  doing is putting together a bunch of puzzle pieces that
  have been pregenerated, and making the simplest, quickest
  decisions they possibly can.~\cite{turner03slashdot}
\end{quote}

Malda's observation points (informally) to the idea of \emph{staging}
the computation performed by a web service.  Indeed, web publishing is
an application area that is naturally staged:\footnote{Nevertheless,
  it seems the literature on meta-programming has not yet acknowledged
  web services~\cite{w3c-wsdl} as a potential application area,
  although \citet{sheard01accomp} does mention \emph{mobile code.}} %
\begin{enumerate}
\item Content (text, images, programs, etc.) created
  off-line is uploaded to the server---the \emph{publish}
  stage;

@@

\screenshot{confcal}{A simple conference calendar web service.}
%, implemented as a
%  PHP script that connects to a PostgreSQL database.  It is not as
%  sophisticated as the service described in the text.}

Now, how might this conference service be staged?  Can
anything be computed off-line (at the \emph{publish} stage)?
Yes!  Since this page is probably part of a much larger site
(whose structure does not change every day), the menus and
other navigation aids can be laid out in advance.  We will
not know which conferences to display until the user
presents some identification (in the form of a `cookie'),
but since the conference data change infrequently, it may
help to prepare the text of each row in advance.

@@

At this point, we should emphasize the importance of \emph{profiling}
in developing scalable web services.  This particular design for the
conference calendar may not be optimal, depending on the number of
entries and the relative speeds of the CPU, memory, database, network,
and disk.  Rather, our aim is to provide a single language in which
the various staging possibilities can be expressed naturally.

This approach is in stark contrast to the status quo, where each
system targets one stage only.  The Website Meta
Language\footnote{\url{http://www.thewml.org/}} is an ``off-line HTML
generation toolkit'' designed for the \emph{publish} stage.  But many
other programs (and countless ad-hoc scripts) spit out HTML pages:
\LaTeX2HTML, for example.  Google reports surprisingly many
programs\footnote{\url{http://google.com/search?q=gedcom+generate+html}}
for creating family tree web sites from genealogy database files;
these also count as publish-stage tools.

The \emph{serve} stage is well-served by the ``server page''
languages, including
JSP,\footnote{\url{http://java.sun.com/products/jsp/}}
ASP,\footnote{\url{http://www.asp.net/}} and
PHP.\footnote{\url{http://www.php.net/}} The Common Gateway Interface
(CGI)\footnote{\url{http://www.w3.org/CGI/}} addresses the
\emph{serve} stage, as do the embedded interpreters (such as
\texttt{mod\_perl} and \texttt{mod\_python} for Apache) that exist to
ameliorate some of the overhead of CGI.

There is, relatively, a paucity of languages that operate client-side
(\emph{display} stage), probably due to the difficulty of securing an
installed base of interpreters.  JavaScript and Java applets are
notable exceptions.

Imagine implementing the conference calendar, as conceived
above, using currently deployed technology: a Perl script
outputs a PHP page which embeds JavaScript!  Values are
passed from one stage to the next as strings, and the
programmer must manage all the quoting and persistence
issues by hand.

Strictly speaking, these languages are not exclusively confined to the
stages that I have indicated: Javascript can be run client-side, PHP
can be run off-line, and so on.  Nevertheless, migrating code between
stages is hard, and the need for quoting and persistence are
practically show-stoppers.  For comparison purposes (further described
in section~\ref{sec:perf}), we staged some code using PHP.  To achieve
persistence of composite variables from one stage to the next, it
contains gems like this:\label{php-gem}
\begin{verbatim}
    <?= "<? \$list = unserialize(\"".
        addcslashes(serialize($list),'"').
        "\"); ?>\n" ?>
\end{verbatim}
where the \texttt{serialize} library function occurs in stage one and
the \texttt{unserialize} in stage two.  Notice that the \verb+$+
preceding the first occurrence of \verb+list+ is quoted, but the
second occurrence is not.  The \texttt{addcslashes} function is needed
in case the serialized representation contains special characters
(such as double quotes or newlines) that would be misinterpreted by
the PHP parser in the next stage.  Interpreting the stage-one program
guarantees nothing about the well-formedness of the stage-two program
(generated as a string).

We present ``\MOC server pages,'' a new domain-specific language for
web applications programming.  It leverages the staging annotations
and static typing of \MOC~\cite{calcagno03meta,taha00metaml} to
provide safe and precise control over the first two stages.  (We leave
further consideration of the display stage as future work.)  The
system is implemented as two components: a translator transforms the
server page language into a \MOC module, which then can be
incorporated into our multi-threaded HTTP/1.1 server (also written in
\MOC).  The scalability gained by staging certain applications is
stunning: in section~\ref{sec:perf} we describe a directory browsing
service where staging yields a factor of 30 improvement in throughput.
The unstaged version would certainly succumb to the Slashdot effect.

The next section sketches the design and translation of \MOC
server pages, and section~\ref{sec:eg} includes some non-trivial
examples.  The server implementation is decribed in
section~\ref{sec:impl}.  Performance and scalability are
discussed in section~\ref{sec:perf}.

@@

The general idea of a \emph{server page} language is that we
  write HTML by default, and embed code %
\cd{\Mquote{}like this\Munquote}.
PHP programmers are familiar with this syntax for embedding code, but
in our case, the code itself is written in \MOC.  Here is a trivial
\MOC server page:
\lgrindfile{first.meta}
and its output:
\begin{quote}
  \textit{Hello,} \textbf{Unhandled exception: Failure("Nice try!")}
\end{quote}
The OCaml function \cd{failwith} raises a \cd{Failure} exception
containing the provided message.  If a code block raises an exception,
the message is sent to the client's browser in bold-face, and the rest
of the page is aborted.

A very common use of code blocks is to print out (i.e., send to the
browser) the result of evaluating some expression.  The syntax
\cd{\Mquote= e\Munquote} is designated for this task; \cd{e} must have
type \cd{string}.  Alternatively, messages may be formatted with
\cd{printf} by placing the format string immediately after the code
delimiter.\footnote{In a departure from the standard OCaml syntax for
  \cd{printf}, the arguments are comma-delimited.  This way, fewer
  parentheses are required when using the escape or lift operators on
  the arguments.}
%%
\lgrindfile{pi.meta}
The output is:
\begin{quote}
  ........ $\pi$ $\div$ 008 is 0.3927
\end{quote}

One more kind of code delimiter is used for \emph{declarations;}
these are lifted to the top of your program, and evaluated during
the publishing stage:
%%
\lgrindfile{perm.meta}
Output:
\begin{quote}
Permissions on current directory are 0755
\end{quote}

Once published, this output will never change!  The \cd{stat} call is
executed only once (because it is in a declaration block), not on each
request.  This is already a rudimentary kind of staging, but with the
annotations of \MOC, we will gain both flexibility and safety, as
we'll see in the rest of this section.

@@

To compute and splice in a regular (non-code) value, we define a
function \cd{{\KWfont let} lift x = \Mbegincode x\Mendcode}; this
takes any value and turns it into code.  We use it like this:
\cd{\Mbegincode 2 * \Msplice(lift(3+4))\Mendcode} The addition is
performed immediately (because it is escaped), and the result is
spliced into the code, producing \cd{\Mbegincode 2 * 7\Mendcode}

@@

\subsection{Translation to \MOC}
To see how all this works, consider how a \MOC server
page is translated into a proper \MOC program, to be executed
at publish time.  Since the program is executed before any
browser has requested the page, it cannot directly return or
output HTML.  Instead, it will construct and return a \cd{code}
object which is subsequently run on each request (serve stage).
The third (display) stage proposed in section~\ref{sec:intro} is not
yet supported by this design.
Figure~\ref{fig:trans.meta} shows a sample \MOC server page,
and figure~\ref{fig:trans.ml} contains its translation.

\mylstf{trans.meta}{This file demonstrates the various
  kinds of code blocks.}

\mylstf{trans.ml}{An automatic translation of the page in
  figure~\ref{fig:trans.meta}.}
%  Declarations have been lifted, and
%  everything else appears within \cd{\Mbegincode brackets\Mendcode}}

Declaration blocks have been lifted to the top; any side effects
contained there are executed when the page is published.  The
constructed code begins on line~5.

The \cd{a b c} represent publish-stage arguments (the names are
specified in the \cd{pragma}), whereas \cd{req} and \cd{puts} are
(fixed) serve-stage arguments.  \cd{req} encapsulates the HTTP request
details, including the headers and query arguments.  \cd{puts} is a
function, provided by the server, to transmit text across the network
to the user's browser.  
%%
The library function \cd{Request.arg} of the type \cd{request\to
  string\to string option} looks up the string value of the request
parameter with a given name.  Since \cd{req} and \cd{puts} are
serve-stage arguments, it is incorrect to use them in the
\emph{publish} (first) stage, and indeed the \MOC type system prevents
this.

A few finer points to note: the translator discards newlines that
immediately follow code blocks (otherwise,
figure~\ref{fig:trans.ml} would be dotted with %
\cd{puts}~\verb+"\n"+ statements).  It automatically appends a
semi-colon or the {\KWfont in} keyword to code blocks, as required
(see, for example, lines 7, 11, and 12 in figure~\ref{fig:trans.ml}.)
%%
Finally, because the translator partially parses the OCaml code
blocks, it is not confused by code delimiters within strings and
comments, nor by other uses of \la{} and \ra{} as operators.

\subsection{Staged code blocks}
Now that we understand how the server page is assembled into a staged
program, the effects of adding \MOC staging operators to our pages
should be predictable.  Below is another example using \cd{stat}, this
time to display the size of the server log file.
With the serve-stage argument \texttt{unit}, we can specify whether
the size should be expressed in bytes (the default), kilobytes, etc.
\lgrindfile{size.meta}%

To stage this, the \cd{stat} and \cd{float\_of\_int} calls could be
lifted into the declaration block, as with the permissions example.
Furthermore, we may also use \cd{lift} and the splice operator to
perform the divisions in advance, even though they are underneath the
{\KWfont match} (which cannot happen until the serve stage):
\lgrindfile{size2.meta}%

Still, the \cd{printf} conversion is performed at serve time, so maybe
it is best to pre-generate that as well:%
\lgrindfile{size3.meta}%

Now, all that remains to execute at serve time is to check the dynamic
\cd{unit} argument and spit out one of three pre-generated strings.
The generated code block looks something like this (substantially
cleaned up from the \MOC pretty-printer, with references to persistent
values resolved in-place):
\lgrindfile{size-out.ml}%

To handle more complex situations, code blocks may be constructed
conditionally and recursively.  Here is an example that
prints a count-down, but removes the loop overhead by expanding to a
sequence of 99 \cd{puts} statements.
\lgrindfile{count.meta}%

The brackets around the \cd{puts} on the last line are necessary,
because \cd{count} splices the \cd{puts} call into the code it
generates.  Omitting the brackets would result in a compile-time type
error.

These tiny examples suggest the ease with which the `boundary' between
the stages can be adjusted, just by tweaking the staging annotations.
Moreover, the static typing of \MOC ensures in advance that our
programs generate type-correct code only.  
%%
The examples also illustrate some of the common uses of the escape
operator, for which we developed syntactic sugar.  The first set of
code blocks we define use the tilde character to indicate that some
part of the enclosed code is escaped:

\begin{tabular}{l@{\quad$\leadsto$\quad}l}
\cd{\Mquote{\Mtilde} a \Munquote} &
\cd{\Mquote{} \Msplice( a ) \Munquote}
\\
\cd{\Mquote{\Mtilde=} b \Munquote} &
\cd{\Mquote{=} \Msplice( b ) \Munquote}
\\
\cd{\Mquote{\Mtilde}{\KWfont let} x = c \Munquote} &
\cd{\Mquote{} {\KWfont let} x = \Msplice( c ) \Munquote}
\\
\cd{\Mquote{\Mtilde}\id{"fmt"} d, e, f\Munquote} &
\cd{\Mquote{}\id{"fmt"} \Msplice(d), \Msplice(e), \Msplice(f)\Munquote}
\\
\end{tabular}

@@

\begin{tabular}{{l@{\quad$\leadsto$\quad}l}}
\cd{\Mquote{\Mbang} a \Munquote} &
\cd{\Mquote{} \Msplice(lift( a )) \Munquote}
\\
\cd{\Mquote{\Mbang=} b \Munquote} &
\cd{\Mquote{=} \Msplice(lift( b )) \Munquote}
\\
\cd{\Mquote{\Mbang}{\KWfont let} x = c \Munquote} &
\cd{\Mquote{} {\KWfont let} x = \Msplice(lift( c )) \Munquote}
\\
\cd{\Mquote{\Mbang}\id{"fmt"} d, e\Munquote} &
\cd{\Mquote{}\id{"fmt"} \Msplice(lift(d)), \Msplice(lift(e))\Munquote}
\\
\end{tabular}

We will see examples of most of these blocks in section~\ref{sec:eg}
and appendix~\ref{sec:dir.meta}.

@@

\MOC server pages support publish-stage arguments, as demonstrated by
the identifiers \cd{a b c} in figures~\ref{fig:trans.meta}
and~\ref{fig:trans.ml}.  The programmer specifies the pattern to be
used with a \cd{\Mquote{}pragma args ...\Munquote} declaration, which
may appear anywhere in the code.

@@

To explore the expressiveness of this design, we now look at a series
of web services, organized to comprise a small web site.\footnote{A
  demo is available at \url{http://comsci.liu.edu/metaserv/}} To give
the services a similar look, we developed a site-wide style sheet, and
an OCaml function to generate a navigation bar from a hierarchical
list of page titles and links.  
%%
\mylst{navspec-sig.ml}{These functions help define the standard site
  layout.}
%%
The page layout functions in figure~\ref{fig:navspec-sig.ml} may be
invoked at any stage.  Naturally, if the page title includes a dynamic
argument, then \cd{preamble} will have to be delayed until the serve
stage.

@@

\screenshot{power127}{Web browser displaying the result of
  7\textsuperscript{127}.}
%  The URI \texttt{/power127} maps to code
%  where the early argument (the exponent) has been set to 127, and the
%  late argument (\texttt{x}, the base) is taken from the query
%  string.}

Judging from its prevalence in the multi-stage programming literature,
we are certain that millions of grateful users would subscribe to an
online service capable of computing the \emph{power} function.  The
screen shot in figure~\ref{fig:power127} illustrates how it works.
The navigation bar shows the exponents for which code has been
pre-generated.\footnote{In the current implementation, it is not
  possible for the user to request other exponents once the server is
  running.  See section~\ref{sec:impl} for an explanation.}  After
selecting the exponent, the user types the base into the form and
presses return.  The result is computed using the arbitrary-precision
\cd{Num} module of the OCaml library.  The complete script appears in
figure~\ref{fig:power.meta}.

\mylstf{power.meta}{The staged power function
 as illustrated in figure~\ref{fig:power127}.}
%  web service.  The exponent \cd{y} is an early (publish stage)
%  argument; but the base \cd{x} is late (serve stage).}

The user's input is converted to a number relatively late in the
script (line 21).  If \cd{num\_of\_string}
generates an exception (perhaps because the user typed non-numeric
text into the box), the navigation bar and form will have already been
output before the error message appears.

In section~\ref{sec:perf}, we will measure the impact of staging on
the scalability of this application.  To derive the un-staged version
for comparison, we simply convert \cd{\Mquote{\Mtilde}} and
\cd{\Mquote{!}} blocks into plain \cd{\Mquote{}} and remove all other
brackets and escapes from the code in figure~\ref{fig:power.meta}.
